Qualitative Probabilistic Programming

### 2015-06-05 paper/paper.tex
fix tables

\documentclass{article}
\usepackage{cite}
\usepackage{listings}
\usepackage{amsmath}
% \usepackage{minted1}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{multicol} \usepackage{fancyheadings} \usepackage{pdfpages}
\usepackage{nips15submit_e,times}
\setlength{\emergencystretch}{10em}

@@

  In the previous example, we that \texttt{randFunction(Categorical(2), Vector(2, Double))} represented 2 axis-aligned Gaussian clusters with equal variances.  In general, \texttt{randFunction} returns a randomized function that is a member of some generalized linear model
    determined by the desired argument types and return type.
    The distribution
    of the function's return value $y$
    is some exponential family whose natural
    parameters are determined from the arguments $x$:

@@

    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for each entry whose row corresponds
    to a sufficient statistics of $y$ that is not a feature and whose column is not 1.
    This ensures that only the
    natural parameters that are features of $y$ are affected by $x$.  The following shows the types
    in Quipp and their corresponding sufficient statistics, features, and English descriptions.

    \begin{tabular}{| c | c | c | c|}
      \hline
      T & $\phi_T(X)$ & $\psi_T(X)$ & Random function class \\

      \hline
    \texttt{Double} & $\begin{bmatrix} X \\ X^2 \end{bmatrix}$ & $\begin{bmatrix} X \end{bmatrix}$ & Linear regression with Gaussian noise \\

      \hline
    \texttt{Categorical(n)} & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $n$-class logistic regression \\

      \hline
    \texttt{Tuple(T1, T2)} & $\begin{bmatrix} \phi_{T1}(X_1) \\ \phi_{T2}(X_2) \end{bmatrix}$
                           & $\begin{bmatrix} \psi_{T1}(X_1) \\ \psi_{T2}(X_2) \end{bmatrix}$ &
      Independent regression of each component
      \\
      \hline
    \end{tabular}

    Note that $\texttt{Vector(n, T)}$ is shorthand for $Tuple(T, T, ..., T)$, with $n$ copies of $T$.

@@

    To infer both latent variables and parameters, we use a Monte Carlo
    expectation maximization algorithm \cite{mcem} on the probabilistic model, iterating stages of
    estimating latent variables using Metropolis Hastings and inferring
    parameters using gradient descent.

@@

    To evaluate performance, for each model, we:
    \begin{itemize}
      \item
        Randomly generate parameters $\theta$
      \item
        Generate datasets $x_{train}, x_{test}$ using $\theta$
      \item
        Estimate $\log P(x_{test} | \theta)$
      \item
        Use the EM algorithm to infer approximate parameters $\hat{\theta}$ from $x_{train}$
      \item
        Estimate $\log P(x_{test} | \hat{\theta})$ and compare to $\log P(x_{test} | \theta)$
    \end{itemize}
    Estimating $\log P(x_{test} | \theta)$ is nontrivial, given that the model contains latent variables.
    We use the Sequential Monte Carlo algorithm for this.  Between observations,

>>>>

\documentclass{article}
\usepackage{cite}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{array}
% \usepackage{minted1}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{multicol} \usepackage{fancyheadings} \usepackage{pdfpages}
\usepackage{nips15submit_e,times}
\setlength{\emergencystretch}{10em}

\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

  In the previous example, we that \texttt{randFunction(Categorical(2), Vector(2, Double))} represented 2 axis-aligned 2-dimensional Gaussian clusters with equal variances.  In general, \texttt{randFunction} returns a randomized function that is a member of some generalized linear model
    determined by the desired argument types and return type.
    The distribution
    of the function's return value $y$
    is some exponential family whose natural
    parameters are determined from the arguments $x$:

@@

    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for each entry whose row corresponds
    to a sufficient statistics of $y$ that is not a feature and whose column is not 1.
    This ensures that only the
    natural parameters that are features of $y$ are affected by $x$.

    Additionally, we provide a $\texttt{randomValue}$ function for convenience, which
    returns a sample from the ``default'' distribution for a given type.  The following shows the types
    in Quipp and their corresponding sufficient statistics, features, and English descriptions.

    \begin{center}
    {\scriptsize
      \begin{tabular}{| c | c | c | c | C{3cm} |}
      \hline
      T & $\phi_T(X)$ & $\psi_T(X)$ & \texttt{randomValue(T)} & Random function class \\

      \hline
    \texttt{Double} & $\begin{bmatrix} X \\ X^2 \end{bmatrix}$ & $\begin{bmatrix} X \end{bmatrix}$ & 
      $\mathcal{N}(0, 1)$ &
      Linear regression\\

      \hline
    \texttt{Categorical(n)} & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & uniform & $n$-class logistic regression \\

      \hline
    \texttt{Tuple(T1, T2)} & $\begin{bmatrix} \phi_{T1}(X_1) \\ \phi_{T2}(X_2) \end{bmatrix}$
                           & $\begin{bmatrix} \psi_{T1}(X_1) \\ \psi_{T2}(X_2) \end{bmatrix}$ &
      \begin{tabular}{l}
        \texttt{[randomValue(T1),}
       \\
     \texttt{ randomValue(T2)]}
      \end{tabular}
                           &
      Independent regression
      \\
      \hline
    \end{tabular}
  }
\end{center}

  Note that \texttt{Vector(n, T)} is shorthand for \texttt{Tuple([T, T, ..., T])} (with $n$ copies of $T$),
  and \texttt{Bool} is shorthand for \texttt{Categorical(2)}.

@@

    To infer both latent variables and parameters, we use a Monte Carlo
    expectation maximization algorithm \cite{mcem} on the probabilistic model, iterating stages of
    estimating latent variables using Metropolis Hastings and inferring
    parameters using gradient descent.  The first iteration uses randomly generated parameters.

@@

    To evaluate performance, for each model, we:
    \begin{itemize}
      \item
        Randomly generate parameters $\theta$
      \item
        Generate datasets $x_{train}, x_{test}$ using $\theta$
      \item
        Estimate $\log P(x_{test} | \theta)$
      \item
        Use the EM algorithm to infer approximate parameters $\hat{\theta}$ from $x_{train}$
      \item
        Estimate the regret $\log P(x_{test} | \hat{\theta}) - \log P(x_{test} | \theta)$
    \end{itemize}
    Estimating $\log P(x_{test} | \theta)$ is nontrivial, given that the model contains latent variables.
    We use the Sequential Monte Carlo algorithm for this, as described in \cite{dippl}.
